Infrastructure Quality and Imperfect Information: Evidence from Piped Water in Manila

### 2021-02-09 phil_paper/paper_booster_pumps_wjv.tex
draft except conclusion

\begin{titlepage} 
%\title{{Borrowing with Unpaid Bills}\thanks{}}
%\title{{Microcredit from Delaying Bill Payments}\thanks{}}
\title{Microcredit from Delaying Bill Payments}
\author{\\[3em]
  William Violette\thanks{Federal Trade Commission, Washington, DC. E-mail: william.j.violette@gmail.com   Any opinions and conclusions expressed herein are those of the author and do not necessarily represent the views of the Federal Trade Commission or its Commissioners.  Many thanks to Matthew Panhans, Miriam Larson-Koester, Adrian Rubli-Ornelas, and Stefano Polloni.} \\
 \\ 
  }
\vspace{30mm}
\date{\vspace{5mm}This Version: \today}
\maketitle
\begin{abstract}

Delaying bill payments to public utilities may provide an important strategy for households with volatile incomes to smooth their consumption.  At the same time, tolerating late payments may reduce net revenues for utilities, which often leads to higher prices to cover costs.  Using billing records from a large water utility in Manila, this paper estimates a household consumption and savings model to evaluate counterfactual payment policies.  A popular proposal to ensure upfront payments --- prepaid metering --- recoups less revenue than is needed to compensate households for their loss of consumption smoothing.  Alternatively, a revenue-neutral policy allowing more late payments increases welfare by encouraging greater consumption smoothing.

\vspace{1in}
\textbf{Keywords:} credit constraints; consumption smoothing; water utilities. \\
\textbf{JEL Codes:} O13; E21; L95. \\
\bigskip
\end{abstract}
\setcounter{page}{0}
\thispagestyle{empty}
\end{titlepage}
\pagebreak \newpage

@@

@@

\section{Data and Setting}

@@

Large investments in booster pumps suggest that households strongly value water pressure and reliability.  Booster pumps are both expensive to purchase and operate. In Manila, booster pumps range in price from 1,200 to 15,000 PhP, which represents a large expense given average monthly household incomes of \input{tables/median_inc}PhP.\footnote{Figures come from scraping results from the first page of searching ``booster pumps'' on the popular online marketplace in Manila, \url{https://www.lazada.com.ph/},  yielding 24 entries with horsepower listed.  Average monthly household income is computed for Metro Manila using the 2015 Family Income and Expenditure Survey.}  The average booster pump uses a 0.9 horsepower engine, which generates monthly costs of around 486 PhP at prevailing energy prices.\footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}   Before pipe replacement, 40\% of households invest in booster pumps that increase water pressure.  After pipe replacement, the share of households using booster pumps drops to 15\%.  This finding is consistent with booster pumps providing an important substitute to pressure from new water pipes.
% = .9*.786*78*8.8
% \footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}

@@

\begin{table}[h!] 
\centering
\caption{Average Survey Responses Before and After Pipe Replacement}\label{table:descriptives}
\vspace{-2mm}
\begin{threeparttable}
\begin{tabular}{@{}l*{1}{KKK}@{}}
\toprule
  & Before & After  & All \\
\midrule
Piped Water Service Quality \\[.5em]
\hspace{1em}$\text{Water has strong pressure (6pm-12am)}^{\dagger}$ & \input{tables/yes_flow_6mid} \\
\hspace{1em}$\text{Water has no pressure (6pm-12am)}^{\dagger}$  & \input{tables/no_flow_6mid} \\
% \hspace{1em}$\text{Hours with pressure per day}^{\dagger}$ & \input{tables/flow_hrs} \\
\hspace{1em}Water interruptions in last 3 months & \input{tables/stop_freq} \\
\hspace{1em}Water has foreign bodies & \input{tables/foreign_bodies} \\
\hspace{1em}Water is discolored & \input{tables/discolored} \\
\hspace{1em}Water has unusual taste/smell & \input{tables/taste_smell} \\
\\[-.5em]
Household Service Quality Investments \\[.5em]
\hspace{1em}Has booster pump & \input{tables/booster} \\
\hspace{1em}Hours booster pump is used per day & \input{tables/booster_use} \\
\hspace{1em}Has water storage tank & \input{tables/drum} \\
\hspace{1em}Has water filter & \input{tables/filter} \\
\hspace{1em}Purchases filtered water & \input{tables/station} \\
\hspace{1em}Purchases from a deepwell & \input{tables/deepwell} \\
\hspace{1em}Spending on non-piped water (PhP) & \input{tables/wrs_exp} \\
\hspace{1em}Drinks from the tap & \input{tables/drink} \\
% \hspace{1em}Boils tap water before drinking & \input{tables/boil} \\
\\[-.5em]
Demographics \\[.5em]
\hspace{1em}Household size & \input{tables/hhsize} \\
\hspace{1em}Employed members & \input{tables/hhemp} \\
\hspace{1em}High-skilled employment & \input{tables/good_job} \\
\hspace{1em}Lives in duplex& \input{tables/sub} \\
\hspace{1em}Lives in single house& \input{tables/single} \\
\hspace{1em}Number of other households sharing tap& \input{tables/SHO} \\
\\[-.5em]
Households & \input{tables/pawsn} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\footnotesize
\item  $\dagger$ when not using booster pump.  Bill, Unpaid Balance, Payment, and Income are in PhP.  Measures exclude months where households remain disconnected through the end of the sample period.  Billing data include households for household-month observations.  Income data include households for household-month observations.  45 PhP = 1 USD \,\,
% See Appendix~\ref{appendix:sampleconstruction} for more details on the sample construction.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{figure}
\begin{center}
\caption{Average Consumption per Household with Years to Pipe Replacement}\label{figure:pipecons}
\includegraphics[scale=1]{tables/pipe_cons.pdf}
\end{center}
%%% add counts and average change in discussion
\end{figure}
% discuss dynamics!!! of figure!!

@@

\begin{table}[h!] % PUT IN DEMOGRAPHICS PLEASE ?!?!?
\centering
\caption{Average Household Characteristics by Prices Charged}\label{table:pricechangestatistics}
\vspace{-2mm}
\begin{threeparttable}
\begin{tabular}{@{}l*{1}{KKK}@{}}
\toprule
  & Always Reg. Price & Always High Price  & Change Reg. to High Price \\
\midrule
Usage per Household (m3)& \input{tables/cv_rs} \\
Household size & \input{tables/hhsize_rs} \\
Employed members & \input{tables/hhemp_rs} \\
High-skilled employment & \input{tables/good_job_rs} \\
Lives in duplex& \input{tables/sub_rs} \\
Lives in single house& \input{tables/single_rs} \\
Other households sharing tap& \input{tables/SHO_rs} \\
\\[-.5em]
Households & \input{tables/N_rs} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\footnotesize
\item  Reg. refers to regular price.

@@

\begin{figure}
\begin{center}
\caption{Usage and Price Changes}\label{figure:usagepricechanges}
\includegraphics[scale=1]{tables/r_to_s_only_graph.pdf}
\end{center}
\end{figure}

\section{Model}

@@

\section{Empirical Strategy}

@@

@@

The remaining term to be estimated in equation (\ref{eq:dvdr}) is the price sensitivity, $\alpha$.  $\gamma_2$ captures the reduced form effect of prices on water usage.  By taking the derivative of equation (\ref{eq:1ststage}) with respect to price, this effect can be expressed as a function of the price sensitivity as well as the effect of price on water quality through endogenous changes in booster pump use, according to the following equation
\begin{align*}
  \gamma_2 = \alpha +  \frac{dQ}{dB} \frac{dB}{dp}
\end{align*}
where $\frac{dQ}{dB}$ is the effect of booster pump use on water quality, which is unobserved.  $\frac{dB}{dp}$ is the effect of prices on booster pump use, which is estimated by $\beta_2$.  If $\frac{dB}{dp}$ is zero, then the expression reduces to $\gamma_2=\alpha$.  Although the booster pump decision is not explicitly modeled, $\frac{dB}{dp}$ may be zero or close to zero when booster pump use is primarily driven by other factors besides the price of water.  For example, booster pumps may be a perfect substitute for water quality such that all households below a certain water quality level choose to use booster pumps while all households above this level do not use booster pumps.  In this example, the booster pump decision is orthogonal to water prices.

Given fixed effects the are identified from

When the outcome is monthly water usage, the estimating equation maps directly onto the equation for optimal household usage given by equation ($\ref{eq:demand}$).  $\beta_1$ captures the effect of pipe replacement on service quality, $\frac{dQ}{dR}$.  Likewise, $\beta_2$ captures the negative of household price sensitivity, $\alpha$.  Remaining fixed preferences for water and levels of service quality are absorbed in household and month fixed effects, $\lambda_i$ and $\theta_t$.  These fixed effects ensure that the effect of pipe replacements is identified from variation in water usage for the same household before and after pipes are fixed while accounting for any temporal usage patterns common to all households.  Likewise, price sensitivity is estimated from variation in price for the same household switching between regular and high prices.  Calendar month fixed effects absorb price variation affecting all households equally over time as the company updates water tariffs.

When the outcome is booster pump use, the estimating equation captures the effect of pipe replacements on booster pump use.  This approach assumes that demand for booster pumps can be linearly approximated by the estimating equation such that $\beta_1$ reflects $\frac{dB}{dR}$.  Since booster pump use is a binary outcome, this approach represents a linear probability model.  Time fixed effects account for variation in booster pump usage over time common to all households while household fixed effects account for differences in booster pump use across households.  $\beta_1$ reflects the effect of pipe replacements on booster pump use under the assumption that no other factors independently affect booster pump use at the same time.  One concern may be that the prices for booster pumps change in response to pipe replacement projects.  Pipe replacements occur in small geographic areas while markets for booster pumps are likely to cover much wider areas.\footnote{Online markets span the entire city.}  Therefore, individual pipe replacement projects are unlikely to influence prices for booster pumps.

% \begin{align}
% \label{eq:1ststage}
% p_{it} &= \phi_1 \text{Post Pipe Rep}_{tl}  +  \phi_2 \text{Post Price Inc}_{it} +  \phi_3\, t\text{ Post Price Inc}_{it} +\phi_4 \,t \,\text{Pre Price Inc}_{it} +  \phi_t  +  \phi_i + \epsilon_{itl} 
% \end{align}
% \begin{align}
% \label{eq:1ststage}
% p_{it} &= \phi_1 \mathbbm{1} \{t\geq \text{Pipe Rep}_{tl}  \} +  \phi_2 \mathbbm{1} \{t\geq \text{High Price}_{tl}  \} +  \phi_3\, t \, \{t\geq \text{High Price}_{tl}  \} +\phi_4 \,t \,\{t< \text{High Price}_{tl}  \} +  \phi_t  +  \phi_i + \epsilon_{itl} 
% \end{align}
% The estimating equation takes the following form
% \begin{align}
% \label{eq:esteq}
% y_{itl} \,=\, \beta_1 \, \text{After Pipe Replacement}_{tl} \,+\, \beta_2 \, p_{it} \, + \, \theta_t \, + \, \lambda_i \, + \epsilon_{itl}
% \end{align}
% where the outcome, $y_{itl}$, is either monthly water usage or booster pump use.  Outcomes are measured for household, $i$, in small area, $l$, at calendar month, $t$. After Pipe $\text{Replacement}_{tl} $ takes a value of one for months after pipe replacement and zero otherwise.  Since the data identify only the year that pipes were replaced, all months from the replacement year onwards are considered after pipe replacement. $p_{it}$ measures the average price faced by each household in each month. $\theta_t$ includes calendar month fixed effects while $\lambda_i$ includes household fixed effects.

>>>>

\begin{titlepage} 
%\title{{Borrowing with Unpaid Bills}\thanks{}}
%\title{{Microcredit from Delaying Bill Payments}\thanks{}}
\title{Infrastructure Quality and Imperfect Information: Evidence from Piped Water in Manila}
\author{\\[3em]
  William Violette\thanks{Federal Trade Commission, Washington, DC. E-mail: william.j.violette@gmail.com   Any opinions and conclusions expressed herein are those of the author and do not necessarily represent the views of the Federal Trade Commission or its Commissioners.  } \\ % Many thanks to Matthew Panhans, Miriam Larson-Koester, Adrian Rubli-Ornelas, and Stefano Polloni.
 \\ 
  }
\vspace{30mm}
\date{\vspace{5mm}This Version: \today}
\maketitle
\begin{abstract}

Public utilities often need to make large investments in infrastructure quality without knowing exactly how much they will benefit consumers.  Using data from a water utility in Manila, this paper estimates the effect of large-scale pipe replacements on consumer surplus both directly through increased water consumption and indirectly through lower spending on private water pumps.  Detailed cost data allow for evaluating counterfactual pipe replacement policies.  Minimizing engineering costs and/or maintaining strict quality standards lead to more infrequent pipe replacements, but small welfare losses relative to a full-information, optimal benchmark.

\vspace{1in}
\textbf{Keywords:} water utility; product quality; imperfect information. \\
\textbf{JEL Codes:} O13; L51; L95. \\
\bigskip
\end{abstract}
\setcounter{page}{0}
\thispagestyle{empty}
\end{titlepage}
\pagebreak \newpage

@@

  % For example, pipe cost, sewer cost, power line costs.  , \cite{frauendorfer2010issues}
% For example, out of the \$472.6 billion USD needed for US water infrastructure in the next 20 years, the bulk of the funding would support large projects such as pipe replacements, water treatment plants, and storage facilities.\footnote{See \cite{epa2015}.}  
% health and productivity 
To ensure service quality, public utilities often require large upfront investments such as replacing water pipes, building power generators, or expanding sewage treatment plants.  The social benefits of these investments are  difficult to measure because investments are infrequent, it can be challenging to precisely locate affected consumers, benefits may be external to consumer demand, and public utilities are often natural monopolies, which means that consumers cannot turn to other providers for better service quality (\cite{spence1975monopoly}).  Without knowing these social benefits, public utilities instead base their investments on meeting service quality benchmarks and/or minimizing engineering cost calculations.  Alternatively, public utility regulators may allow private companies to determine and profit from quality investments if companies are better able to measure the benefits and costs of investments.  It remains an empirical question which regulatory policies best ensure socially efficient levels of infrastructure quality.

Quality regulation may have particularly large welfare consequences in developing cities where aging infrastructure and underinvestment contribute to persistently low levels of service quality.  For example, over 20\% of the piped water supplied in developing cities leaks from pipes before reaching consumers, resulting in low water pressure and intrusion of contaminants (\cite{liemberger2006challenge}).  Longterm quality deficits have led households to invest in private substitutes for service quality.  In Mexico City, 45\% of households use rooftop storage tanks to deal with intermittent water supply (\cite{baisa2010welfare}) while households in urban India turn to a combination of private water tanks, pumps, and/or tube wells (\cite{mckenzie2009urban}).  While difficult to measure, spending on quality substitutes may be an important part of the welfare costs of low service quality.  
% Recent research in developing countries also links improvements in infrastructure quality to external benefits such as improved health\footnote{See \cite{bhalotra2017urban}, \cite{galiani2005water}, \cite{ashraf2017water}.} and productivity.\footnote{See \cite{allcott2016electricity} and \cite{hardy}.}

This paper estimates the effect of water pipe replacements on consumer welfare in Manila, Philippines and then evaluates alternative pipe replacement policies.  Like many developing cities, Manila initially suffered from low levels of quality investment by the public utility, which was privatized through a public-private parternship in 1997 (\cite{dumol2000manila}).  Unlike in many developing cities, the public-private partnership launched a massive infrastructure upgrading program around 2008, which replaced at least 50\% of the existing pipes by 2015.  This paper analyzes billing and location records from 1.5 million water connections over this period as well as cost, pipe location, and survey data.  Analyzing the impacts of these investments may provide useful lessons for other developing cities weighing similar programs.

The empirical strategy leverages the staggered roll-out of pipe replacements to identify its effects on household water consumption and electric booster pump use --- a common substitute for low water pressure.  The estimation strategy recovers the causal effect of replacements from pre/post changes in outcomes compared to changes in areas without replacements under the assumption that areas would have evolved similarly in the absence of replacements.  This assumption is supported by the fact that engineers organized the pipe replacement program to meet hydrological constraints, which were unlikely to have been related to trends in local water demand. % This assumption is further supported by evidence of relatively flat trends in outcomes leading up to replacement projects.

Translating consumption changes into consumer surplus also requires estimating the price elasticity of demand.  The empirical strategy analyzes novel price variation where some households are suddenly switched to 22\% higher prices at different times.  These switches occur when the water company notices any business activity at households' residences, which usually entails small, roadside food stands.  This event-study approach complements the largely cross-sectional and time-series approaches to estimating price elasticities in the literature.\footnote{See \cite{szabo2015value} and \cite{mcrae2015infrastructure} as well as \cite{olmstead2009reduced} for a review.}

Results indicate that pipe replacements have large effects on consumer welfare driven in particular by lower spending on quality substitutes.  Pipe replacements are associated with a 10\% increase in household water consumption and a 30 percentage point decline in booster pump use.  Guided by a simple model of household water demand, these estimates suggest economically large increases in monthly consumer surplus on the order of 175 PhP or 35\% of an average household water bill.  Consumer utility from greater consumption and from savings on booster pumps both contribute roughly equally to the total increase in consumer surplus, highlighting the importance of private quality substitutes in evaluating infrastructure projects.  Gains in consumer surplus also appear concentrated among lower socioeconomic status households.  % (68+107)/500

The paper calculates the longterm welfare consequences of replacing pipes more or less frequently in different areas of the city.  To calculate longterm welfare, the paper assumes that pipes deteriorate linearly over time and leverages how the observed pipe replacements effectively reset the age of the pipes from 25 years old to zero.  Therefore, the estimated effects of pipe replacements on revenue, leakage costs, and consumer surplus summarize how welfare evolves over time.  Estimates are conducted separately for 11 different areas of the city to simulate the spatially targeted nature of infrastructure investments.  % according to maximize different criteria

The paper then compares welfare and pipe replacement frequencies under imperfect information to those under a full-information benchmark holding fixed all other aspects of water provision.  In the first case, the government regulator only observes the age of the pipes and chooses to replace pipes every 25 years, which matches the observed experience in Manila.  By replacing pipes much less frequently, this policy lowers monthly welfare by around 20\% of an average water bill compared to the full-information benchmark.  This result suggests that Manila would have experienced substantial welfare gains by replacing its pipes sooner.

In the second case, the regulator observes the water pressure reaching consumers and sets a minimum pressure threshold, replacing pipes when pressure falls below this threshold.  While the pressure threshold outperforms 25-year replacements, welfare is very sensitive to the exact threshold levels.  The third case allows the regulator to observe and minimize pipe replacement and leakage costs, further improving welfare by replacing pipes more frequently on average.  However, monthly welfare still lags behind the full-information benchmark by an amount equal to 6\% of an average water bill.

The last case allows a regulated, private firm to schedule pipe replacements in order to maximize profits (ie. revenues minus costs) under the assumption that the firm is able to perfectly and freely observe demand and costs.  Since prices in Manila are set above marginal costs, extra consumption generated by pipe replacements would give the firm an additional incentive to replace pipes quickly.  This context emulates ``price-cap'' regulation where the regulator sets a maximum price to cover operating costs and the firm is able to profit from additional investments.\footnote{See \cite{joskow2007regulation}.}  Profit maximization results in the next most frequent pipe replacements aside from the full-information benchmark.  As a result, monthly welfare lags behind the full-information benchmark by an amount equal to only 2\% of an average water bill.  Consumer welfare is only slightly smaller than total welfare because high pipe replacement costs limit any potential profits earned by the firm.

This paper builds on a long literature on public utility regulation in industrial organization.  While previous research, pioneered by \cite{laffont1993theory}, largely focuses on information asymmetries between firms and regulators (\cite{mcrae2015infrastructure}, \cite{lim2018dynamic}), this paper examines the case where both the regulator and the firm have imperfect information, which may be especially relevant to infrastructure quality.  This paper also provides an empirical examination of the regulatory challenges inherent in measuring and regulating quality outlined by \cite{spence1975monopoly}.

This paper also contributes to a recent development literature studying the impacts of improved public utility service quality.  Recent research has mainly focused on the impact of public utilities on health (\cite{bhalotra2017urban}, \cite{kosec2014child}, \cite{ashraf2017water}), education (\cite{zhang2016long}), and  productivity (\cite{allcott2016electricity}, \cite{hardy}).  This paper measures the direct consumer surplus from improved water service quality in a similar way to how \cite{mcrae2015infrastructure} recovers demand for continuous electricity service.  Like \cite{baisa2010welfare}, this paper also studies the welfare implications of spending on private substitutes to water service quality.

The paper proceeds with a discussion of the context in Manila and the datasets in Section~\ref{section:data}.  Section~\ref{section:descriptives} provides descriptive statistics on water use in Manila.  Section~\ref{section:model} develops a simple model of water demand and Section~\ref{section:empiricalstrategy} outlines the empirical strategy to estimate the model.  Section~\ref{section:results} discusses the results.  Section~\ref{section:counter} incorporates the results into counterfactual policy simulations and Section~\ref{section:conclusion} concludes.

% \cite{mcrae2015infrastructure} considers the value of power outages to consumers but not . 
% \cite{mcrae2015infrastructure}
% \cite{laffont2005regulation}
% DEVELOPMENT CONTRIBUTION mostly focuses on externalities; this one is 
% The paper also abstracts away from potential health or productivity externalities associated with water quality improvements.

% The counterfactual results are likely to be most relevant to urban settings with little investment in pipe maintenance, rapid development, and little health or productivity externalities from water use.  In Manila, pipe replacements have large impacts on consumer surplus .  These large impacts likely stem from 
% The counterfactuals predict very short replacement rates of 8 to 12 years compared to an engineering literature in the US that predicts at least 50 year pipe replacement rates.  
% Since the counterfactuals rely on estimated pipe replacement effects, the counterfactual results assume that

% EXTERNAL VALIDITY
% These results are only valid for the levels of maintenance during the public utility, especially while the public utility was there...
% relative results may be instructive

% Costs of gathering the information ! 
% External validity broadly !  when and when not...

PUT FOOTNOTES IN TO EXPLAIN THE AMOUNTS!!

  % The counterfactuals then choose the optimal pipe replacement rates to satisfy their welfare criteria for each area.

% The \textit{full-information} counterfactual provides an optimal benchmark where the regulator chooses when to replace pipes to maximize total welfare given full information on demand and costs.  In the \textit{regular-replacement} counterfactual, the regulator simply replaces all pipes at a fixed interval, which loosely follows the approach taken in Manila.  In the \textit{quality-standards} counterfactual, the regulator establishes a minimum water pressure threshold and replaces pipes when pressure falls below the threshold.  In the \textit{cost-minimization} counterfactual, the regulator schedules replacements to minimize leakage costs over time as well as upfront replacement costs.  In the \textit{firm-level} counterfactual, the regulator lets the firm decide when to replace pipes under the assumption that the firm has full demand and cost information.

%%% goal: by 1, first draft intro; goal by evening: first draft conclusion

Conclusion should discuss REGULATION THEORY!!

% The government regulator gives the water company discretion to assign households to a high-price tariff schedule if they have any business activity at their residence, which provides a novel source of price variation in the context of public utilities.  In the vast majority of cases, the high price tariff is applied to households that operate small food stands (or ``Sari-Sari'' stores).

% Public utilities cope with this uncertainty by delaying investments,  and/or relying on engineering 
% quality is provided in a monopoly setting where consumers face few alternatives.  
% first because there's a monopoly.
% Since the welfare benefits of these investments are often difficult to measure, 
% Information is really hard to get for water pipes...; degrade over time; leakage detection systems are costly, etc.

@@

\section{Data and Setting}\label{section:data}

@@

Large investments in booster pumps suggest that households strongly value water pressure and reliability.  Booster pumps are both expensive to purchase and operate. In Manila, booster pumps range in price from 1,200 to 15,000 PhP, which represents a large expense given average monthly household incomes of \input{tables/median_inc}PhP.\footnote{Figures come from scraping results from the first page of searching ``booster pumps'' on the popular online marketplace in Manila, \url{https://www.lazada.com.ph/},  yielding 24 entries with horsepower listed.  Average monthly household income is computed for Metro Manila using the 2015 Family Income and Expenditure Survey.}  The average booster pump uses a 0.9 horsepower engine, which generates monthly costs of around 486 PhP at prevailing energy prices.\footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}   Before pipe replacement, 40\% of households invest in booster pumps that increase water pressure.  After pipe replacement, the share of households using booster pumps drops to 15\%.  This finding is consistent with booster pumps providing an important substitute to pressure from new water pipes.

% = .9*.786*78*8.8
% \footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}

@@

\begin{table}[h!] 
\centering
\caption{Average Survey Responses Before and After Pipe Replacement}\label{table:descriptives}
\vspace{-2mm}
\begin{threeparttable}
\begin{tabular}{@{}l*{1}{KKK}@{}}
\toprule
  & Before & After  & All \\
\midrule
Piped Water Service Quality \\[.5em]
\hspace{1em}$\text{Water has strong pressure (6pm-12am)}^{\dagger}$ & \input{tables/yes_flow_6mid} \\
\hspace{1em}$\text{Water has no pressure (6pm-12am)}^{\dagger}$  & \input{tables/no_flow_6mid} \\
% \hspace{1em}$\text{Hours with pressure per day}^{\dagger}$ & \input{tables/flow_hrs} \\
\hspace{1em}Water interruptions in last 3 months & \input{tables/stop_freq} \\
\hspace{1em}Water has foreign bodies & \input{tables/foreign_bodies} \\
\hspace{1em}Water is discolored & \input{tables/discolored} \\
\hspace{1em}Water has unusual taste/smell & \input{tables/taste_smell} \\
\\[-.5em]
Household Service Quality Investments \\[.5em]
\hspace{1em}Has booster pump & \input{tables/booster} \\
\hspace{1em}Hours booster pump is used per day & \input{tables/booster_use} \\
\hspace{1em}Has water storage tank & \input{tables/drum} \\
\hspace{1em}Has water filter & \input{tables/filter} \\
\hspace{1em}Purchases filtered water & \input{tables/station} \\
\hspace{1em}Purchases from a deepwell & \input{tables/deepwell} \\
\hspace{1em}Spending on non-piped water (PhP) & \input{tables/wrs_exp} \\
\hspace{1em}Drinks from the tap & \input{tables/drink} \\
% \hspace{1em}Boils tap water before drinking & \input{tables/boil} \\
\\[-.5em]
Demographics \\[.5em]
\hspace{1em}Household size & \input{tables/hhsize} \\
\hspace{1em}Employed members & \input{tables/hhemp} \\
\hspace{1em}Respondent has high-skilled employment & \input{tables/good_job} \\
\hspace{1em}Lives in duplex& \input{tables/sub} \\
\hspace{1em}Lives in single house& \input{tables/single} \\
\hspace{1em}Number of other households sharing tap& \input{tables/SHO} \\
\\[-.5em]
Households & \input{tables/pawsn} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\footnotesize
\item  $\dagger$ when not using booster pump.  Bill, Unpaid Balance, Payment, and Income are in PhP.  Measures exclude months where households remain disconnected through the end of the sample period.  Billing data include households for household-month observations.  Income data include households for household-month observations.  45 PhP = 1 USD \,\,
% See Appendix~\ref{appendix:sampleconstruction} for more details on the sample construction.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{figure}
\begin{center}
\caption{Average Water Usage per Household with Years to Pipe Replacement}\label{figure:pipecons}
\includegraphics[scale=1]{tables/pipe_cons.pdf}
\end{center}
%%% add counts and average change in discussion
\end{figure}
% discuss dynamics!!! of figure!!

@@

\begin{table}[h!] % PUT IN DEMOGRAPHICS PLEASE ?!?!?
\centering
\caption{Average Household Characteristics by Prices Charged}\label{table:pricechangestatistics}
\vspace{-2mm}
\begin{threeparttable}
\begin{tabular}{@{}l*{1}{KKK}@{}}
\toprule
  & Always Reg. Price & Always High Price  & Change Reg. to High Price \\
\midrule
Usage per household (m3)& \input{tables/cv_rs} \\
Household size & \input{tables/hhsize_rs} \\
Employed members & \input{tables/hhemp_rs} \\
Respondent has high-skill employment & \input{tables/good_job_rs} \\
Lives in a duplex& \input{tables/sub_rs} \\
Lives in a single house& \input{tables/single_rs} \\
Other households sharing tap& \input{tables/SHO_rs} \\
\\[-.5em]
Households & \input{tables/N_rs} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\footnotesize
\item  Reg. refers to regular price.

@@

\begin{figure}
\begin{center}
\caption{Water Usage and Price Changes}\label{figure:usagepricechanges}
\includegraphics[scale=1]{tables/r_to_s_only_graph.pdf}
\end{center}
\end{figure}

\section{Model}\label{section:model}

@@

\section{Empirical Strategy}\label{section:empiricalstrategy}

@@

To estimate the effects of pipe replacement change over time, additional specifications replace the terms, $\gamma_1  \text{Post Pipe}_{tl} $ and $\beta_1  \text{Post Pipe}_{tl} $, with the terms, $\sum_{k=0}^{K} \gamma_1^{k}  \text{Post Pipe}_{tl}^{k} $ and  $\sum_{k=0}^{K} \beta_1^{k}  \text{Post Pipe}_{tl}^{k} $, where $k$ indexes years after pipe replacement, $\text{Post Pipe}_{tl}^{k}$ is an indicator variable for each $k$ year after pipe replacement, and coefficients, $\gamma_1^{k}$ and $\beta_1^{k}$, measure the lasting effects at each $k$ year after pipe replacement.

To estimate heterogeneity according to household characteristics, further specifications replace the terms, $\gamma_1  \text{Post Pipe}_{tl} $ and $\beta_1  \text{Post Pipe}_{tl} $, with the terms, $\sum_{j=1}^{J} \gamma_1^{j} \mathbbm{1}\{D^{j}=1\} \times \text{Post Pipe}_{tl} $ and  $\sum_{j=1}^{J} \beta_1^{j} \mathbbm{1}\{D^{j}=1\} \times \text{Post Pipe}_{tl} $, where $j$ indexes household characteristics, $\mathbbm{1}\{D^{j}=1\}$ is an indicator variable for households with characteristic, $j$, and coefficients, $\gamma_1^{j}$ and $\beta_1^{j}$, measure the effects of pipe replacements for households with characteristic, $j$.

@@

The remaining term to be estimated in equation (\ref{eq:dvdr}) is the price sensitivity, $\alpha$.  $\gamma_2$ captures the reduced form effect of prices on water usage.  By taking the derivative of equation (\ref{eq:1ststage}) with respect to price, this effect can be expressed as a function of the price sensitivity as well as the effect of price on water quality through endogenous changes in booster pump use, according to the following equation
\begin{align}
\label{eq:psens}
  \gamma_2 = \alpha +  \frac{dQ}{dB} \frac{dB}{dp}
\end{align}
where $\frac{dQ}{dB}$ is the effect of booster pump use on water quality, which is unobserved.  $\frac{dB}{dp}$ is the effect of prices on booster pump use, which is estimated by $\beta_2$.  If $\frac{dB}{dp}$ is zero, then the expression reduces to $\gamma_2=\alpha$.  Although the booster pump decision is not explicitly modeled, $\frac{dB}{dp}$ may be zero or close to zero when booster pump use is primarily driven by other factors besides the price of water.  For example, booster pumps may be a perfect substitute for water quality such that all households below a certain water quality level choose to use booster pumps while all households above this level do not use booster pumps.  In this example, the booster pump decision is orthogonal to water prices.

\section{Results}\label{section:results}
% Columns (1) and (3) provide average effects while Columns (2) and (4) provide separate effects for each year after pipe replacements.

Table~\ref{table:mainregs} includes the main estimates of the effects of average price and pipe replacements on water and booster pump use.  The first row of Column (1) finds an average increase in household water usage of 2 m3 per month after pipe replacement.  The estimate is statistically significant at the 1\% level and economically large, representing a 10\% increase in average household water use.  The first row of Column (2) finds that pipe replacement projects are associated with a 22 percentage point average decrease in booster pump use, which is statistically significant at the 1\% level.  This estimate also represents an economically significant decrease given that the average level of booster pump use across the full sample is 16\%.

Columns (2) and (4) provide separate effects for each year after pipe replacements.  According to Column (2), the effect on water use is highest in the first two years after pipe replacement at around 2.3 m3 before declining linearly to around 2 m3 at 6 years after pipe replacement.  Column (4) finds a similar declining pattern for booster pump use between one and two years after pipe replacement.  Booster pump effects beyond two years are unavailable because the connection survey only covers a limited time period.  Columns (2) and (4) find a relatively smaller effect in the year of construction likely because many projects were completed at some point during this year.

The declining effect over time is consistent with pipes gradually deteriorating.  The linear trend in the decline matches civil engineering research documenting a roughly linear relationship between pipe failure and pipe age especially within the first 50 years after installation.\footnote{See \cite{ward2017deterioration}, \cite{kleiner2001comprehensive}, and \cite{aydogdu2015estimation}.}  Extrapolating the linear trend suggests that households would return to their pre-replacement consumption levels within around 37 years.  In contrast, the average age of the replaced pipes is around 25 years, which suggests that the new pipes may be more durable or subject to better maintenance than the replaced pipes.

In terms of estimating price sensitivity, the second row of column (1) provides a reduced form effect of prices on water use of 0.59, which is statistically significant at the 1\% level.\footnote{Appendix Table~\ref{table:regstages} indicates that the first-stage is statistically significant and the t-statistic is at least 17 across specifications.}  Since the second row of column (2) finds zero effect of price on booster pump use,\footnote{Although the price effect is identified from only the 208 connections that are surveyed after being switched into a high price tariff, the estimate is relatively precise ruling out effects above 3.7\% and below -2\% with 95\% confidence.} then the reduced form effect of price on usage can be interpreted directly as household price sensitivity according to equation (\ref{eq:psens}).  Given an average price of 21.5 PhP/m3, this estimate implies a price elasticity of 0.64.  This elasticity estimate is in the middle of the range from 0.01 to 0.98 in the developing world.\footnote{See \cite{szabo2015value}, \cite{diakite2009proposal}, and \cite{strand2005water}.}

These results allow for benchmarking the average effect of pipe replacements on consumer welfare over 6 years according to equation (\ref{eq:dvdr}).  Using inputs from column (1) of Table~\ref{table:mainregs}, consumer surplus from improved water quality computes to around 68 PhP per household/month while consumer surplus from lower booster pump usage computes to around 107 PhP per household/month.  This result suggests that both factors are similarly important in driving changes in consumer welfare.

% is equal to the product of average water use and the effect of pipe replacements on water use weighted by the price sensitivity,
